---
title: "Case Study 2: University tuition"
author: "Kelly Bodwin"
output: html_document
---

## Get the data

We will be attempting to find a linear regression that models college tuition rates, based on a dataset from US News and World Report.  Alas, this data is from 1995, so it is very outdated; still, we will see what we can learn from it.

*****
### Question 1:

a) The dataset is located at `http://kbodwin.web.unc.edu/files/2016/09/tuition_final.csv`; figure out how to use the code you were given last time for `read.csv( )` and `read.table( )` to read the data into **R** and call it `tuition`. Use the functions we learned last time to familiarize yourself with the data in `tuition`. 

```{r, eval = TRUE}
  # YOUR CODE HERE
tuition = read.csv("http://kbodwin.web.unc.edu/files/2016/09/tuition_final.csv")



```

b) Make a new variable in `tuition` called `Acc.Rate` that contains the acceptance rate for each university.

```{r, eval = TRUE}
  # YOUR CODE HERE
tuition$Acc.Rate<- tuition$Accepted/tuition$Applied

```

c) Find and print the row of the data that corresponds to UNC ("University of North Carolina at Chapel Hill").

```{r, eval = TRUE}
  # YOUR CODE HERE
tuition[tuition$Name == "University of North Carolina at Chapel Hill",]
```



## Question 7:

a. The variable `Spending` contains the expenditure of the school per student. Suppose we want to make a regression that predicts tuition cost from the expenditure per student.  Make a linear model for `Spending` versus `Out.Tuition`.  Comment on the summary of the model, and plot it on an appropriate scatter plot. Does the model seem to be a good fit for this data?
```{r, eval = TRUE}
  # YOUR CODE HERE

lm_svt = lm(Out.Tuition ~ Spending , data = tuition)


plot(tuition$Spending, tuition$Out.Tuition, main = "Scatterplot of Spending vs. Out.Tuition", xlab = "Spending", ylab = "Out.Tuition", pch = 4, cex = .75, col = "brown")
abline(lm_svt, col = "black")

summary(lm_svt)

  
```

```
based on the r-squared value and the scatterplot, the linear model doesn't seems to fit the data very well and needs to be adjusted(.4409). 
```

b. Plot the residuals versus the values of `Spending`.  Do you notice any issues? *Hint: Use your own function `predict_yval()` or the built-in function `predict(my_lm)`.  You will need to think about the problem of missing data (`NA`s).*

```{r, eval = TRUE}
  # YOUR CODE HERE
new_data<- na.omit(tuition, cols=c("Spending", "Out.Tuition"))
new_lm <- lm(Out.Tuition~Spending, data =new_data)
new_data$predicted_y = predict(new_lm)
new_data$residual <- new_data$Out.Tuition - new_data$predicted_y

plot(new_data$Spending,new_data$residual,pch=4,bg="red",col="royal blue", main="Residual of Out Tuition Based on Spending", xlab="Spending",ylab="Residual")
abline(0,0,lw= 1.5)

```

```
The residual plot is not very symmetrical, as most points are clustered to the left with a some points to the right. The plot looks pretty random but the few points on the right are causing problems in the data.
```


c. Use `plot()` to look at the automatic diagnostics.  What is each plot showing? What seems to be going wrong here?  Which schools are marked as outliers?

```{r, eval = TRUE}
  # YOUR CODE HERE
  plot(new_lm)
```

```
917(Antioch),496(Johns Hopkins),976(Franklin and Marshall George) are the schools marked as outliers. There are clear outliers that are affecting the shape of the residual plot; this is also seen in the qq plot with outliers deviating far from the line in the qq plot. 
```

d. Roughly speaking, an outlier is "influential" if it changes the regression line when you remove it.  Decide for yourself which data points are influential outliers. Recalculate the linear model without any outliers, and plot it on a scatterplot.

```{r, eval = TRUE}
  # YOUR CODE HERE
  outlier_rm = new_data[-c(917,496,976),]
  plot = plot(new_data$Spending,new_data$Out.Tuition,pch=4,col="brown")
  rm_out= lm(Out.Tuition~Spending, data = new_data)
  summary(rm_out)
  abline(rm_out, lw=2)
```

***
### Question 8:
a. Now suppose we want to make a regression that predicts tuition cost from the size of the student body.  Make a linear model for `Size` versus `Out.Tuition`.  Comment on the summary of the model, and plot it on an appropriate scatter plot, and use `plot( )` to look at the diagnostics.  Does the model seem to be a good fit for this data?
```{r, eval = TRUE}
  # YOUR CODE HERE
  sizevout = lm(Out.Tuition~Size, data=tuition)
  summary(sizevout)
  plot(tuition$Size,tuition$Out.Tuition, xlab="Size",ylab="Out.Tuition",main="Scatterplot of Size vs. Out.Tuition")
  plot(sizevout)

```

```
based on the scatterplot, the points seem random with no linear shape, which hints that a linear regression is not a good fit. Furthermore the r-squared value is almost 0, which is another strong indication that the the model is not a good fit.The QQ plot is not linear as well, which is yet another indication that a linear model may not be the best fit for the data. 
```

b. Remake your scatterplot, this time including the option `col = tuition$Public`.  What did this change?  Can you use this information to explain why the regression line in (a) did not fit well?
```{r, eval = TRUE}
  # YOUR CODE HERE
 plot = plot(tuition$Spending,tuition$Out.Tuition,pch=4,col= tuition$Public, main="Spending v. Out.Tuition with Public Differentiated",xlab="Spending",ylab="Out.Tuition")

```

```
the plot now differientiates be ween spendings associated with private and public schools. Since the disparity between tuition costs for public and private universities is quite large, the correlation between the data will seem much lower. It is difficult to correlate the spendings to tuition costs when the costs and schools are vastly different. Private and public spendings will both have less significance if both are included in the same dataset.  
```

c. Make separate linear regressions of `Size` versus `Out.Tuition` for private schools and for public schools.  Plot both of these, appropriately colored, on your scatterplot.  Comment on the output and diagnostics.
```{r, eval = TRUE}
  # YOUR CODE HERE
  private = tuition[tuition$Public == 2,]
  lm_private = lm(Out.Tuition~Size, data=private)
  summary(lm_private)
  public = tuition[tuition$Public == 1,]
  lm_public = lm(Out.Tuition~Size, data=public)
  summary(lm_public)
  
  
```

```
Overall, most of the statistics increased, including F-Statistic, t-value and p-value, and the Adjusted r-Squared for both private and public schools. While the r-squared is stil low, the F-Stastic still suggestst that one group is affecting the outcome.        
```
***

## Multiple Linear Regression

We have seen that a college's tuition relates to its size, its spending per student, and its average SAT score.  We have also seen that this relationship may change based on whether the school is public or private.  Ideally, instead of making separate regressions for each relationship, we could combine them all into a multiple regression. Fortunately, **R** makes this easy with `lm()`.

***
### Question 9:

a) Run the following code to perform a multiple regression.  Interpret the results.

```{r, eval = TRUE}
  mult.1 <- lm(Out.Tuition ~ Size + Avg.SAT + Avg.ACT + Spending + Acc.Rate, data = tuition)
  
  summary(mult.1)
```

```
the very higher f-statstic of 137.6 is one indication that at least one of the predictor values explains the Tuition Rate. Size and Spending both have a large effect on the response variable jduging by the large t-value and very small p-value. Acc.Rate and Avg.Sat have a much smaller effect on the data. Observing the Adjusted R-squared value, only 59.34% of the out.tuition can be explained by the predicted values, therefore the predictors are moderately correlated to the response.  
```

b) We can also mix and match continuous variables with categorical ones.  Let's add `Public` to the regression.  The following two models are slightly different, but give essentially identical output.  What is the difference between them, and why is it important even though the output still the same?

```{r, eval = TRUE}
  mult.2 <- lm(Out.Tuition ~ Size + Avg.SAT + Avg.ACT + Spending + Public, data = tuition)
  mult.3 <- lm(Out.Tuition ~ Size + Avg.SAT + Avg.ACT + Spending + factor(Public), data = tuition)
  summary(mult.2)
  summary(mult.3)


```

```
The intercept values that were calculated were lower when the categorical public(-9.316e+03) was added in comparison to the factor(Public)(-5.96e+03). In terms of the linear regression equation, it makes a big difference since the y-value at 0 is even lower than before, affecting the equation of the line. 
```

c) It is still important to check diagnostics in a multiple regression, although it can be harder to track down the source of problems.  Use `plot( )` to look at diagnostics for `mult.3`, and comment.

```{r, eval = TRUE}
  # YOUR CODE HERE
  plot(mult.3)
```

```
The residual plot and QQ plot still shows that there are 3 outliers that are affecting the graphs. The residual plot looks better than the original residual plotted earlier, as the points look more random and symmetrical excluding the outliers. The QQ plot also looks very straight, all which overall hint/indicate that a linear model is a good fit.   
```

***
### Question 10:
a) A big problem in multiple regression is *collinearity*, which means that two or more explanatory variables are correlated with each other. Read the documentation for `pairs( )`, and then use it on the variables involved in `mult.3`.  *Hint:  You can use the option `col = tuition$Public` in `pairs( )`*

```{r, eval = FALSE}
  # YOUR CODE HERE

  pairs( ~ Size + Avg.SAT + Avg.ACT + Spending + factor(Public), data = tuition,lower.panel=NULL,col = tuition$Public)
```

b) Do any of the variables seem strongly related?  What is their correlation?

```{r, eval = FALSE}
  # YOUR CODE HERE
  summary(lm(Avg.SAT~Avg.ACT, data=tuition))
  summary(lm(Out.Tuition~Avg.ACT,data=tuition))
```

```
based on the pairs and the r-squared of Avg.Sat and Avg.ACT, those two variables seems strongly correlated with each other
```

c) Explain in your own words why the correlation between the variables you discussed in (a) could be a problem.

```
having correlation within data causes alot of effects on the data. One big effect though is if two predictor values are correlated, one will lose relation to the respons variable. FOr example, the t-value for Avg.Act and Out.Tuition is highly significant without Avg.SAT in the summary, but when Avg.SAT is included, ACT becomes insignificant in relation to Out.Tuition.  
```
***

## Sneak Preview: Interaction Terms

We saw in 12c that whether a school is public or private can affect not only the tuition, but also how the tuition relates to other variables.  In a multiple regression, this effect can be captured through interaction terms, which are expressed by `var1:var2`, and measure how much one variable changes the effect of the other.  

Read the following paragraph from the documentation `?formula` for some shortcuts for including interactions:
```
In addition to + and :, a number of other operators are useful in model formulae. The * operator denotes factor crossing: a*b interpreted as a+b+a:b. The ^ operator indicates crossing to the specified degree. For example (a+b+c)^2 is identical to (a+b+c)*(a+b+c) which in turn expands to a formula containing the main effects for a, b and c together with their second-order interactions. The %in% operator indicates that the terms on its left are nested within those on the right. For example a + b %in% a expands to the formula a + a:b. The - operator removes the specified terms, so that (a+b+c)^2 - a:b is identical to a + b + c + b:c + a:c. It can also used to remove the intercept term: when fitting a linear model y ~ x - 1 specifies a line through the origin. A model with no intercept can be also specified as y ~ x + 0 or y ~ 0 + x.
```
***
### Question 11:
Create your own multiple regression that predicts tuition from whichever variables you choose, as well as some interaction terms between `Public` and other variables.  Don't worry about using any official methods to pick variables; simply try a few things and choose the model that seems best.  Interpret the results; in particular, think very carefully about what the coefficient for an interaction term with `Public` might mean.

```{r, eval = TRUE}
  # YOUR CODE HERE
mult.own <- lm(formula = Out.Tuition ~ Accepted + Avg.SAT + Spending:factor(Public), data = tuition)

summary(mult.own)
plot(mult.own)



```

```
The multiple linear regression chosen has 3 predictors: Accepted, Avg.SAT and Spending which is dependent on whether the school is private or public. Observing the plots, the QQ plot is roughly linear but a few points do not fall on the line. In addtion this residual is not perfectly symmetrical and has outliers skewing the plot. However, the F-Stastic is very high(340.2) therefore the p-value is <2.2e-16, which suggests that one or more of the predictor values are affecting the Out.Tuition. The predictor value seem to be a good estimate of the linear model since the R-Squared value is .6477, which is much higher than the previous R-Squared values calculated previously. Observing each of the predictor values,the p-values are all very significant, with values almost close to 0. Depending on whether the school was private or public, the beta coefficient is either positive(public schools) and negative(private schools), suggesting that as spending increases for public schools, the out.tuition is lower, whereas private schools experience an increase in out.tuition. The standard errors for each of the predictor are all pretty small, which suggests a more accurate spread for each of the factors. 


```
***