---
title: 'Case Study 1: College Basketball'
author: "prepared by Kelly Bodwin"
output: html_document
---

In this Case Study, you will refresh your memory of STOR 155 while you learn some basic commands and tools for analyzing data with **R**.  We'll be looking at some data from college basketball games in 2015.

Run the following **R** code to load the data into your RStudio and take a look at it.

## Summarizing data

```{r, eval = TRUE}
# Load dataset
bball = read.csv("http://kbodwin.web.unc.edu/files/2016/06/basketball.csv")

# Look at dataset
head(bball)
summary(bball)

```

The command `read.csv( )` will read a dataset into R from your computer or from online.  "csv" stands for "comma separated value", a common file type where the data is listed in a text file, with variables separated by commas.  For now, you don't need to worry about the details of `read.csv( )`.  Once you have loaded the data, the command `summary( )` will tell you about the variables in the dataset and their values.  Another useful function is `head( )`, which shows you the first 6 rows of the dataset.

***

### Question 1:  

a) Look at the outputs of `summary(bball)` and `head(bball)`, and describe the variables using vocabulary from STOR 155.  
```
```

b) If `head( )` shows the first 6 rows of the dataset, what command do you think might show the *last* 6 rows?  Try out your proposed function and see what happens
```
```

c) Try the commands `ncol( )` and `nrow( )`.  What do these do?  How could you get the same information from `head( )`, `summary( )`, and/or the command you figured out in part (b)?
```
```

***

Sometimes, we will want to look at individual entries, rows, or columns of our data matrix.  We can do this using brackets `[ ]` after our dataset.  We can also look at a variables (columns) by name using the *$* symbol.  Try the following examples.

```{r, eval = FALSE}
# Look at a single row
bball[123, ]

# Look at a single column
bball[ , 5]
bball$Team.Score

# Look at a single entry
bball[123, 5]
bball$Team.Score[123]

# Calculate mean, median, variance, and standard deviation
mean(bball$Team.Score)
median(bball$Team.Score)
var(bball$Team.Score)
sd(bball$Team.Score)

```

***

### Question 2:

a) What is the difference between `mean(bball$Team.Score)` and `mean(bball[,5])`?  Why might it be useful to have two ways to get access the variable `Team.Score`?
```
```

b) In plain English, what were the events of the game represented by the first row of the dataset?
```
```
*(Note:  If you don't know much about basketball - for example, if you don't know what it means to play a game "Home" versus "Away" - ask people around you.)*


***


All these commands we have been using, like `summary( )` and `mean( )` are called *functions*.  A function can take all different kinds of input depending on what you are trying to do: datasets, vectors such as `bball$Team.Score`, etc. An important skill in **R** is figuring out for yourself how functions work.

For example, type `?boxplot` into your **R** console.  A help page will pop up telling you about this function.  Notice that under **Usage**, it says `boxplot(x, ...)`.  This tells you that you need to supply something called *x* to the function, and the rest of the input is optional.  But what is *x*?  Ah-ha!  There is a section called **Arguments**, which tells us that *x* is the vector of values you want to put in a boxplot.

Run the code below to make a boxplot of the team scores of college basketball games.

```{r, eval = FALSE}

# make boxplot of team scores
boxplot(bball$Team.Score)

```


***

### Question 3: 

a) Now check out `?hist`, a function for making histograms.  Below is basic code to make a histogram of `Team.Scores`, and also code for the same histogram but with a lot of the optional input changed.  Mess around with these inputs until you understand what each is doing.   

```{r, eval = FALSE}

# Boring histogram
hist(bball$Team.Score)

# Fancy histogram
hist(bball$Team.Score, breaks = 5, main = "I am a title", xlab = "I am an x-axis label", col = "grey", freq = FALSE)

```
Explain in your own words what `breaks` and `freq` change about the histogram.
```
```

b) The optional inputs `main`, `xlab`, `ylab`, and `col` are common to most plotting functions.  Use what you learned in (a) to make a boxplot of `Team.Scores` with proper axis labels and title.

```{r, eval = FALSE}

# make boxplot of team scores
boxplot(bball$Team.Score, YOUR CODE HERE)

```

c) To check if the histogram is Normal, or to help visualize its shape, we might want to overlay a Normal curve on top of the histogram.  The code below will do so - but the curve doesn't fit very well.  
```{r, eval = FALSE}

# Boring histogram
hist(bball$Team.Score, freq = FALSE)

# overlay Normal Curve

curve(dnorm(x, mean=120, sd=20), 
      add = TRUE, col = "blue", lwd = 2)
```
Explain what the role is of the functions `curve( )` and `dnorm( )`.  Why did we put `add = TRUE` in the inputs?
```
```

d) Alternatively, we can overlay a line that is a "smoothed" version of the data, as follows:

```{r, eval = FALSE}

# overlay smoothed curve
lines(density(bball$Team.Score),
      col = "red", lwd = 2, lty = 2)

```

What is the difference between `lines( )` and `curve( )`?  When might we want to use `density( )`, and when would it be better to overlay a Normal curve on a histogram?
```
```



e)  Now make your own histogram with well-chosen inputs and with a Normal overlay that fits better.  Would you say the data looks Normal?

```{r, eval = FALSE}
YOUR CODE HERE
```

***

## Subsetting

One of the most powerful qualities of **R** is the ability to select a subset of a dataset. Suppose we want to look only at games involving UNC or Duke.  We would need to figure out which rows of `bball` involve one of those teams, and then make a new dataset out of only those rows.  

For this, we will use *booleans*, which are variables with the value `TRUE` or `FALSE`.  Play around with the following code until you feel comfortable with `==`, `>`, `<`, and `%in%` as well as `&` (and) and `|` (or).

```{r, eval = FALSE}
# booleans practice

1 == 1
1 == 2
1 < 2

1 == 1 | 1  > 2
1 == 1 & 1 > 2

```

You can make up your own vector using the function `c( )`, which stands for "concatenate".  This is like making a new variable - the variable can contain anything you want, such as numbers, strings, booleans. Try the example below to make a vector and subset it. Note that we can use either `<-` or `=` to store information in a variable.

```{r, eval = FALSE}

vec <- c("cat", "dog", "horned toad", "Her Majesty Queen Elizabeth", "dog")
vec

# Some more booleans
vec == "dog"
"dog" == vec
vec %in% c("dog", "cat")
c("dog", "cat") %in% vec


# Finding indices

which(vec == "dog")
which(vec %in% c("dog", "cat"))
which(c("dog", "cat") %in% vec)

# Subsetting
new = vec[vec %in% c("dog", "cat")]
new

```

***

### Question 4:

a) The following code will give you an error.  What happened?
```{r, eval = FALSE}

vec = c(1, 2, 3, "4")
vec + 2

```
```
```

b) The following code will NOT give you an error?  What is going on here?
```{r, eval = FALSE}

vec = c(TRUE, FALSE, FALSE, TRUE)
vec + 2

```
```
```

c) Now we are ready to make a new dataset.  We'll get a list of booleans to tell us where UNC or Duke's games are, and use that to subset the datset `bball`.

Try running each of the following lines of code.  None of them will make the datset we want.  What was the problem with each one?

```{r, eval = FALSE}

# Make new dataset with only UNC or Duke games


#A
my_subset = bball[Team == "North Carolina" | Team == "Duke", ]

#B
my_subset = bball[bball$Team == "North Carolina", bball$Team == "Duke"]

#C
my_subset = bball[bball$Team = "North Carolina" | bball$Team = "Duke", ]

#D
my_subset = bball[bball$Team == "North Carolina" & bball$Team == "Duke", ]

#E
unc_games = which(bball$Team == "North Carolina")
my_subset = bball[unc_games | bball$Team == "Duke", ]

#F
my_subset = bball[bball$Team == "North Carolina" | bball$Team == "Duke"]

```

```

```


d) Now write your own code to make the correct dataset.

```{r, eval = FALSE}
YOUR CODE HERE

```

***

## Z-Scores and t-scores

Alright, enough of that data wrangling.  Time to do some statistics.

Check out `?Normal`.  These are some functions that will help us calculate probabilities about the Normal distribution. (No more using Table A!)  The most important ones are `pnorm` and `qnorm`.

`pnorm(q)` will tell you the probability of a standard Normal being below the value `q`

`qnorm(p)` will tell you the z-score that has area `p` below it on a standard Normal curve

***

### Question 5

a) For each of the following lines of code, think about what the result will be **before** running the code.  **Draw a picture for each one to visualize what is going on with `pnorm` and `qnorm`.**

```{r, eval = FALSE}
# practice with Normal densities in R

#i
pnorm(0)
qnorm(0)

#ii
pnorm(100)
qnorm(100)

#iii
qnorm(pnorm(0))
qnorm(pnorm(7))

#iv
pnorm(qnorm(0))
pnorm(qnorm(0.5))

#v
pnorm(0, sd = 10)
pnorm(0, mean = 1, sd = 10)

#vi
qnorm(0.05)
qnorm(0.05, sd = 10)
qnorm(0.05, mean = 1, sd = 10)

```

b) Why did you get an error in part (ii)?
```
```

***

Now use this code to make a new variable for the total score of a game:

```{r, eval = FALSE}
# Make new variable
bball$Total.Score = bball$Team.Score + bball$Opponent.Score
```

We will use *z-scores* and *t-scores* to think about whether a game is unusually high scoring.

***

### Question 6:

a) As you may have noticed, the dataset `bball` actually displays each game twice: once for each team.  Make a new dataset with each game listed only once by subsetting `bball`.
```{r, eval = FALSE}
YOUR CODE HERE
```

b) On Feb 17, 2016, UNC played Duke.  Using the Normal distribution, what percent of games have higher scores than the UNC/Duke game?  (Assume that the mean and standard deviation of `Team.Score` are actually the *population* mean and standard deviation.)  
```{r, eval = FALSE}
YOUR CODE HERE
```

c) What percentage of games in the dataset did we observe to be higher scoring than the UNC/Duke game?  The functions `sum( )` and `length( )` will help you answer this question.

```{r, eval = FALSE}
YOUR CODE HERE
```

d)  What is the difference between what we did in (b) and (c)?  Do you think the Normal approximation is reasonable for this data?  Why or why not?
```
```

***

Recall that *t-scores* are used instead of *z-scores* when the population standard deviation is unknown.  The functions `pt` and `qt` work almost same way as `pnorm` and `qnorm`, but for the t-distribution instead of the Normal.  However, be careful, and read `?pt` for help!  These functions don't let you enter the mean and standard deviation as input - you need to figure out what do about that!

***

### Question 7:
Use all your new **R** skills to answer this question: Was the Feb 17th game between UNC and Duke particularly high scoring *for a UNC game*?

```{r, eval = FALSE}
YOUR CODE HERE

```

***

## Confidence Intervals and Proportions

You now have all the **R** knowledge you need to make some confidence intervals!  You may wish to go over your lecture notes for this section, especially to remind yourself how to deal with proportions.

***

### Question 8:

a) Make a 95% confidence interval for the number of points UNC scores in a given game.  You will need to think about which **R** commands will give you critical values of the *t*-distribution, and how to use these to make a confidence interval.

```{r, eval = FALSE}
YOUR CODE HERE

```

b) What percentage of games did UNC win in 2015-2016?  Make a 95% confidence interval for their win percentage.


```{r, eval = FALSE}
YOUR CODE HERE
```

***

## Hypothesis Testing

You have now had lots of practice learning to use a function by reading the documentation.  Part of the point of this course is for you to become familiar enough with **R** to learn new commands and functions without being shown how to use them.  This will make you a skillful (and hireable!) programmer in the future.

Check out `?t.test` and `?prop.test`.  Figure out what these functions do, what input they take, etc. Then answer the following questions.

***

### Question 9:

a) Does UNC tend to win more games than they lose?  That is, is there evidence at the 0.05 level that the "true" probability of UNC winning a given game in 2015-2016 is larger than 0.5?

```{r, eval = FALSE}


```

b) Based on how many points they tend to score in a game, would you say UNC and Yale were equally good teams? 

```{r, eval = FALSE}

```
```

```

c) Based on win percentage, would you say UNC and Yale were equally good teams?  Discuss this result and the result in (b).

```{r, eval = FALSE}

yale.won = sum(yale_games$Team.Result == "Win")
n2 = nrow(yale_games)

prop.test(c(unc.won, yale.won), c(n, n2))

```
```
```

***
## Comparing multiple means (Analysis of Variance)

What if we want to compare more than one team? In lecture, you learned about using an Analysis of Variance (ANOVA) F-test to check if more than two means are equal.  We will use the function `aov( )` to find out if the big three North Carolina teams - UNC, Duke, and NC State - all tend to score the same number of points.

***
### Question 10
a) Make a dataset called `nc_games` that includes only games for the North Carolina teams, and then alter the code below to create a box plot of the scores for the three North Carolina teams.  Does it look like any of the means are significantly different?

```{r, eval = FALSE}


```

b) Perform an ANOVA F-test on the means.  Interpret the output.  Is there evidence that the average scores of the three teams are not all equal?

```{r, eval = FALSE}

```

```
```

***