# Matrices in R
**Created by John Palowitch, UNC Chapel Hill**

**Edited by Jan Hannig, UNC Chapel Hill**


### Conceptual Questions 
Please read the document ***Matrices in R.docx*** before completing this assignment. This assignment has two goals:

-Learning new commands in R from a written source.

-Using R for Matrix Calculations
<br><br>
1. Suppose:
\[
\mathbf{A} = \begin{bmatrix}2&3&4&5\\-1&0&3&2\end{bmatrix},\;\;\;\;\mathbf{B} = \begin{bmatrix}4&-1&10&0\\-2&-2&5&6\end{bmatrix}
\]
\[
\mathbf{a} = \begin{bmatrix}6&2&-11&1\end{bmatrix},\;\;\;\;\mathbf{b} = \begin{bmatrix}4\\2\\3\\0\end{bmatrix},\;\;\;\;\mathbf{c} = \begin{bmatrix}-3\\-10\end{bmatrix}
\]

    a. Write down the dimensions of the following (you can use a comment box that you create). If the operation is not possible, write "N.P."
  
        i. $\mathbf{A}^T\mathbf{B}$  4X4 (first one done for you!)
        ii. $\mathbf{B}\mathbf{A}^T$ 2X2  
        iii. $\mathbf{A}\mathbf{B}$  N.P  
        iv. $\mathbf{A}^T\mathbf{A}$ 4x4
        v. $\mathbf{b}^T\mathbf{A}$  N.P.
        vi. $\mathbf{a}\mathbf{A}^T$ 1X2
        vii. $\mathbf{B}\mathbf{b}$  2X1
        viii. $\mathbf{B}\mathbf{c}^T$  N.P.
        ix. $\mathbf{a}\mathbf{c}$    N.P.
        x. $\mathbf{b}\mathbf{a}$    4X4
        
    
        
    b. Again, write down the dimensions of the following (on this sheet, next to the expression). If the operation is not possible, write "N.P."
    
        i. $(\mathbf{A}^T\mathbf{B})^{-1}\mathbf{a}$ N.P.
        ii. $\mathbf{c}(\mathbf{B}\mathbf{A}^T)^{-1}$ N.P.
        iii. $(\mathbf{B}^T\mathbf{B})^{-1}\mathbf{B}^T\mathbf{c}$ N.P.
        iv. $\mathbf{b}^T\mathbf{A}^T\mathbf{c}$ 1X1
        v. $(\mathbf{B}\mathbf{B}^T)^{-1}\mathbf{B}\mathbf{b}$ 2X1
        
### Applied Questions 
2. In a code block below, create and print the following matrices and vectors:
\[
\mathbf{A} = \begin{bmatrix}-6&0&1\\1&10&7.5\end{bmatrix},\;\;\;\;\mathbf{B} = \begin{bmatrix}-15&1&1\\5.25&1&2.5\end{bmatrix}
\]
\[
\mathbf{a} = \begin{bmatrix}6&2&1\end{bmatrix},\;\;\;\;\mathbf{b} = \begin{bmatrix}2\\-7\end{bmatrix}
\]

  
```{r, eval=TRUE}   
   A <- matrix(c(-6,0,1,1,10,7.5),nrow=2,byrow=TRUE)
   B <- matrix(c(-15,1,1,5.25,1,2.5),nrow=2,byrow=TRUE)
   a<-matrix(c(6,2,1),nrow=1,byrow=TRUE)
   b<-matrix(c(2,-7),ncol=1)
   
   A
   B
   a
   b
```



3. Compute (with **R**, in a new code block below) and print the following products:
    a. $\mathbf{A}^T\mathbf{B}$
    b. $\mathbf{B}\mathbf{a}^T$
    c. $\mathbf{b}^T\mathbf{A}$
    d. $\mathbf{A}^T\mathbf{b}$
    e. $\mathbf{a}^T\mathbf{a}$
    f. $\mathbf{a}\mathbf{a}^T$
    g. $(\mathbf{A}^T\mathbf{A})^{-1}$
    h. $(\mathbf{A}^T\mathbf{A})^{-1}\mathbf{A}^T\mathbf{b}$
    i. $\mathbf{b}^T\mathbf{A}\mathbf{a}^T$
    j. $\mathbf{a}\mathbf{B}\mathbf{A}^T\mathbf{b}$

#a.
```{r, eval=TRUE} 
At=t(A)
QA = At%*%B
QA
```


#b.
```{r, eval=TRUE} 
QB = B%*%t(a)
QB
```


#c.
```{r, eval=TRUE} 
QC=t(b)%*%A
QC
```


#d.
```{r, eval=TRUE} 
QD = t(A)%*%b
QD
```


#e.
```{r, eval=TRUE} 

QE = t(a)%*%a
QE
```

#f.
```{r, eval=TRUE} 

QF = a%*%t(a)
QF

```

#g. 

QG= solve((t(A)%*%A))

Determinant is 0 zero, so the inverse cannot be calculated because inverse is 1/determinant





#h.

Determinant is zero, so the inverse cannot be calculated because inverse is 1/determinant
```{r, eval=TRUE}

#solve((t(A)%*%A))%*%t(A)%*%b


```

#i.
```{r,eval=TRUE}
#i
QI= t(b)%*%A%*%t(a)
QI
```


#j.
a is a 1X3 matrix whereas B is a 2X3, therefore multiplication cannot take place since the dimensions would not line up
```{r, eval=TRUE} 
#QJ=a%*%B

#%*%t(A)%*%b
 
```
```   
   
   
   


