---
title: 'Case Study 1: College Basketball'
author: "prepared by Kelly Bodwin"
output: html_document
---

In this Case Study, you will refresh your memory of STOR 155 while you learn some basic commands and tools for analyzing data with **R**.  We'll be looking at some data from college basketball games last year.

Run the following **R** code to load the data into your RStudio and take a look at it.

## Summarizing data

```{r, eval = TRUE}
# Load dataset
bball = read.csv("http://kbodwin.web.unc.edu/files/2016/06/basketball.csv")

# Look at dataset
head(bball)
#summary(bball)
ncol(bball)
nrow(bball)

```

The command `read.csv( )` will read a dataset into R from your computer or from online.  "csv" stands for "comma separated value", a common file type where the data is listed in a text file, with variables separated by commas.  For now, you don't need to worry about the details of `read.csv( )`.  Once you have loaded the data, the command `summary( )` will tell you about the variables in the dataset and their values.  Another useful function is `head( )`, which shows you the first 6 rows of the dataset.

***



### Question 6:

a) As you may have noticed, the dataset `bball` actually displays each game twice: once for each team.  Make a new dataset with each game listed only once by subsetting `bball`.
```{r, eval = TRUE}

no_repeat = bball[bball$Team.Result == "Win",]



```

b) On Feb 17, 2016, UNC played Duke.  Using the Normal distribution, what percent of games have higher scores than the UNC/Duke game?  (Assume that the mean and standard deviation of `Team.Score` are actually the *population* mean and standard deviation.)  
```{r, eval = FALSE}
unc = bball[bball$Date == "2/17/16" & bball$Team == "North Carolina", ]

unc$Total.Score

mean_games = mean(no_repeat$Total.Score)
sd_games = sd(no_repeat$Total.Score)

pnorm(unc$Total.Score,mean_games,sd_games, lower.tail =FALSE)
```

c) What percentage of games in the dataset did we observe to be higher scoring than the UNC/Duke game?  The functions `sum( )` and `length( )` will help you answer this question.

```{r, eval = FALSE}
total_ppl= length(no_repeat$Total.Score)
score_high= sum(unc$Total.Score > no_repeat$Total.Score)
percent = 1- (score_high/total_ppl)
percent
```

d)  What is the difference between what we did in (b) and (c)?  Do you think the Normal approximation is reasonable for this data?  Why or why not?
```
b uses the cumulative function and takes a summation of all the probabilites above the sample mean to calculate the percentage of scores higher than that game whereas c goes through the code to actually find how many games unc/duke scored higher than everyone else and divides by the total number of all games played subtracted by 1 to get a percent value. 

```

***

Recall that *t-scores* are used instead of *z-scores* when the population standard deviation is unknown.  The functions `pt` and `qt` work almost same way as `pnorm` and `qnorm`, but for the t-distribution instead of the Normal.  However, be careful, and read `?pt` for help!  These functions don't let you enter the mean and standard deviation as input - you need to figure out what do about that!

***

### Question 7:
Use all your new **R** skills to answer this question: Was the Feb 17th game between UNC and Duke particularly high scoring *for a UNC game*?

```{r, eval = FALSE}

unc_subset = bball[bball$Team == "North Carolina",]
unc_mean = mean(unc_subset$Total.Score)
unc_sd = sd(unc_subset$Total.Score)

unc_pnorm = pnorm(unc$Total.Score,unc_mean,unc_sd,lower.tail = FALSE)
unc_pnorm

"Based on the pnorm value, this game was not a particularly high scoring game since the 62.23% games were higher than the score of that UNC/Duke game"
```

***

## Confidence Intervals and Proportions

You now have all the **R** knowledge you need to make some confidence intervals!  You may wish to go over your lecture notes for this section, especially to remind yourself how to deal with proportions.

***

### Question 8:

a) Make a 95% confidence interval for the number of points UNC scores in a given game.  You will need to think about which **R** commands will give you critical values of the *t*-distribution, and how to use these to make a confidence interval.

```{r, eval = TRUE}
unc_subset = bball[bball$Team == "North Carolina",]


unc_ind_sd = sd(unc_subset$Team.Score)
unc_ind_mean = mean(unc_subset$Team.Score)


t_star = qt(.975, df = length(unc_subset) -1,)

error = t_star*(unc_ind_sd)/sqrt(length(unc_subset))


lower_unc = unc_ind_mean - error
upper_unc = unc_ind_mean + error 

lower_unc
upper_unc

```

b) What percentage of games did UNC win in 2015-2016?  Make a 95% confidence interval for their win percentage.


```{r, eval = TRUE}
z = qt(.975,df = length(unc_subset)-1,)

unc_win = length(which(unc_subset$Team.Result == "Win"))
total_unc = length(unc_subset$Team.Result)
unc_percent = unc_win/total_unc

MOE = sqrt((unc_percent*(1-unc_percent))/unc_win)*z

upper_percent = unc_percent+MOE
lower_percent = unc_percent-MOE

lower_percent
upper_percent

```
***

## Hypothesis Testing

You have now had lots of practice learning to use a function by reading the documentation.  Part of the point of this course is for you to become familiar enough with **R** to learn new commands and functions without being shown how to use them.  This will make you a skillful (and hireable!) programmer in the future.

Check out `?t.test` and `?prop.test`.  Figure out what these functions do, what input they take, etc. Then answer the following questions.

***

### Question 9:

a) Does UNC tend to win more games than they lose?  That is, is there evidence at the 0.05 level that the "true" probability of UNC winning a given game in 2015-2016 is larger than 0.5?

```{r, eval = TRUE}
unc_win = length(which(unc_subset$Team.Result == "Win"))
total_unc = length(unc_subset$Team.Result)

prop.test(unc_win,total_unc,p=.5,alternative="greater",correct=FALSE)

"Based on the p-value of 8.066e-05, the value is significant enough to reject the null-hypothesis. Therefore UNC tended to win more games than they lost"


```

b) Based on how many points they tend to score in a game, would you say UNC and Yale were equally good teams? 

```{r, eval = TRUE}
unc_games = bball[bball$Team == "North Carolina",]
yale_games = bball[bball$Team == "Yale",]
unc_scores = unc_games$Team.Score
yale_scores = yale_games$Team.Score

t.test(unc_scores,yale_scores, alternative = "two.sided",var.equal =FALSE)

"based on the p-value of 0.01611, there is evidence that the p-value is signifcant enough to conclude that one team is better than the other
based on scoring"
```
```

```

c) Based on win percentage, would you say UNC and Yale were equally good teams?  Discuss this result and the result in (b).

```{r, eval = TRUE}
yale_games = bball[bball$Team == "Yale",]
yale.won = sum(yale_games$Team.Result == "Win")
unc.won = sum(unc_subset$Team.Result == "Win")
n = nrow(unc_subset)
n2 = nrow(yale_games)

prop.test(c(unc.won, yale.won), c(n, n2))


"Both teams are equally good teams, as the p-value of .9585 is too high to reject the null. Cannot determine if one team is better than the other"
```
```
```

***
## Comparing multiple means (Analysis of Variance)

What if we want to compare more than one team? In lecture, you learned about using an Analysis of Variance (ANOVA) F-test to check if more than two means are equal.  We will use the function `aov( )` to find out if the big three North Carolina teams - UNC, Duke, and NC State - all tend to score the same number of points.

***
### Question 10
a) Make a dataset called `nc_games` that includes only games for the North Carolina teams, and then alter the code below to create a box plot of the scores for the three North Carolina teams.  Does it look like any of the means are significantly different?

```{r, eval = TRUE}
nc_games = bball[bball$Team =="North Carolina State",]
unc_games = bball[bball$Team == "North Carolina",]
duke_games = bball[bball$Team == "Duke",]
names <- c("North Carolina State", "North Carolina", "Duke")

boxplot(nc_games$Team.Score,unc_games$Team.Score,duke_games$Team.Score, names = names,xlab ="Teams",ylab = "Scores", main = "Box plot of North Carolina Team's Scores", col = c("red", "lightblue","blue"))


"Based on the boxplots, the Mean score of NC State is much lower than UNC and Duke's Scores. UNC and Duke have similar mean scores but UNC is slightly higher"




```

b) Perform an ANOVA F-test on the means.  Interpret the output.  Is there evidence that the average scores of the three teams are not all equal?

```{r, eval = TRUE}
all_teams = bball[bball$Team =="North Carolina State" | bball$Team == "North Carolina" | bball$Team == "Duke",]
anova=aov(all_teams$Team.Score~all_teams$Team)

summary(anova)


"Based on the p-value of the data(0.0284), the p-value is significant enough to reject the null. We can then conclude that the average scores of at least one of the three teams are not equal"

```

```
```

***